---
title: "pml_p_g_a"
author: "Shridhar Benni"
date: "22/10/2020"
output: html_document
---

## Background  
These days there are all kinds of wearable devices like Fitbit, Apple watch, Nike watch etc. It is now possible to collcet large amount of data about physical activities.
In this project we will get some insights from these data of the candidates by doing some analyses.

## Data Sources  
The training data for this project is available at this url:  
[https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv](https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv)  
The test data is available at this url:  
[https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv](https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv)  

## Intended Results  
The goal of this project is to predict the manner in which the people did the exercise using all the data from these wearables. 

## Reproducibility  
We will be using soem packages, and will be setting some global variables.

```{r warning=FALSE, error=FALSE}
library(rattle)
library(caret)
library(rpart)
library(rpart.plot)
library(corrplot)
library(randomForest)
library(RColorBrewer)
```  
Finally, load the seed using following code:  
```{r warning=FALSE, error=FALSE}
set.seed(56879)
```  

Download the dataset to the `data` folder :

```{r warning=FALSE, error=FALSE}
trainLink <-"https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
testLink <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
trainCsv <- "./data/pml-training.csv"
testCsv  <- "./data/pml-testing.csv"
if (!file.exists("./data")) {
  dir.create("./data")
}
if (!file.exists(trainCsv)) {
  download.file(trainLink, destfile = trainCsv, method = "curl")
}
if (!file.exists(testCsv)) {
  download.file(testLink, destfile = testCsv, method = "curl")
}
rm(trainLink)
rm(testLink)
```  

## Reading Data  
Reading the two csv files into the dat frame.

```{r warning=FALSE, error=FALSE}
trainData <- read.csv(trainCsv)
testData <- read.csv(testCsv)
dim(trainData)
dim(testData)
rm(trainCsv)
rm(testCsv)
```  
The training data contains `r dim(trainData)[1]` observations and `r dim(trainData)[2]` variables, while the testing data contains `r dim(testData)[1]` observations and `r dim(testData)[2]` variables.
We have to predict `classe` variable.

## Cleaning Data  
Cleaning the data set and getting rid of the missing values.

1. Clean the <b>Near Zero Variance</b> Variables from the datset.  
```{r warning=FALSE, error=FALSE}
val <- nearZeroVar(trainData, saveMetrics = TRUE)
head(val, 20)
train01 <- trainData[, !val$nzv]
test01 <- testData[, !val$nzv]
dim(train01)
dim(test01)
rm(trainData)
rm(testData)
rm(val)
```  

2. Removing some columns of the dataset that do not contribute to our analysis.
```{r warning=FALSE, error=FALSE}
reg <- grepl("^X|timestamp|user_name", names(train01))
train <- train01[, !reg]
test <- test01[, !reg]
rm(reg)
rm(train01)
rm(test01)
dim(train)
dim(test)
```  

3. Removing columns that contain `NA's` values. 
```{r warning=FALSE, error=FALSE}
col <- (colSums(is.na(train)) == 0)
train <- train[, col]
test <- test[, col]
rm(col)
```  

The cleaned training data contains `r dim(train)[1]` observations and `r dim(train)[2]` variables, while the testing data contains `r dim(test)[1]` observations and `r dim(test)[2]` variables.  

Correlation Matrix of Columns in the Training Data set is. 
```{r warning=FALSE, error=FALSE}
corrplot(cor(train[, -length(names(train))]), method = "color", tl.cex = 0.5)
```  

## Partitioning Training Set  
We split the cleaned training data into 2 parts with 70% as training set and 30% as validation set.
```{r warning=FALSE, error=FALSE}
set.seed(5679) # For reproducibile purpose
inTrain <- createDataPartition(train$classe, p = 0.70, list = FALSE)
valid <- train[-inTrain, ]
train <- train[inTrain, ]
rm(inTrain)
```  
The Data now consists of `r dim(train)[2]` variables with the observations divided as following:  
1. Training Data: `r dim(train)[1]` observations.  
2. Validation Data: `r dim(valid)[1]` observations.  
3. Testing Data: `r dim(test)[1]` observations.  

## Data Modelling  

### Decision Tree  
We use <b>Decision Tree</b> algo to create a model.
```{r warning=FALSE, error=FALSE}
treeModel <- rpart(classe ~ ., data = train, method = "class")
prp(treeModel)
```  

Estimating the performance of the model on the <b>validation</b> data.
```{r warning=FALSE, error=FALSE}
treePredict <- predict(treeModel, valid, type = "class")
confusionMatrix(valid$classe, treePredict)
accuracy <- postResample(treePredict, valid$classe)
os <- 1 - as.numeric(confusionMatrix(valid$classe, treePredict)$overall[1])
rm(treePredict)
rm(treeModel)
```  

The Estimated Accuracy of the Random Forest Model is `r accuracy[1]*100`% and the Estimated Out-of-Sample Error is `r os*100`%.  

### Random Forest
We use a <b>Random Forest</b> predictive model  for activity recognition.
```{r warning=FALSE, error=FALSE}
rfModel <- train(classe ~ ., data = train, method = "rf", trControl = trainControl(method = "cv", 5), ntree = 250)
rfModel
```  

Now, we estimate the performance of the model on the <b>validation</b> data set.  
```{r warning=FALSE, error=FALSE}
rfPredict <- predict(rfModel, valid)
confusionMatrix(valid$classe, rfPredict)
accuracy <- postResample(rfPredict, valid$classe)
os <- 1 - as.numeric(confusionMatrix(valid$classe, rfPredict)$overall[1])
rm(rfPredict)
```  

The Estimated Accuracy of the Random Forest Model is `r accuracy[1]*100`% and the Estimated Out-of-Sample Error is `r os*100`%.  
Random Forests yielded better Results, as expected!  

## Predicting The Manner of Exercise for Test Data Set  
```{r warning=FALSE, error=FALSE}
rm(accuracy)
rm(os)
predict(rfModel, test[, -length(names(test))])
```  

## Generating Files 
```{r warning=FALSE, error=FALSE}
write_files = function(x){
  n = length(x)
  for(i in 1:n){
    filename = paste0("./problem_id_",i,".txt")
    write.table(x[i], file = filename, quote = FALSE, row.names = FALSE, col.names = FALSE)
  }
}
```  

```{r warning=FALSE, error=FALSE}
write_files(predict(rfModel, test[, -length(names(test))]))
rm(rfModel)
rm(train)
rm(test)
rm(valid)
rm(write_files)
```  